---
title: "New small synthetic data"
author: "Sergio Picart-Armada"
date: "2nd March, 2020"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      error = FALSE, warning = FALSE, 
                      fig.height = 4)
```

## Loading packages

```{r cars}
library(plyr)
library(dplyr)
library(tidyr)
library(tibble)
library(magrittr)

library(SummarizedExperiment)

library(ggplot2)

library(pcaMethods)
library(data.table)

theme_set(theme_bw())

if (!dir.exists("02_data")) dir.create("02_data")

set.seed(1)
```

Generate data with:

* Signatures: 
  - 3 organisms (mmu, rno, hsa)
  - 6 signatures
  - Metadata: treatment/animal model

* Pathways
  - 6 pathways
  - 3 types (signal, mixed, noise) 
  - Another metadata column
  
* Reference data
  - Quantitative y: -2 to 2
  - 8 patients per group
  - 300 genes: 150 with signal, 150 with noise
  - Signal: mean(x_i) = mean(y_i), by class. Noise: mean(x_i) = 0

## Orthology mapping

```{r}
devtools::load_all()

df.mouse2human <- orthIDcon("mmusculus", "hsapiens")
df.rat2human <- orthIDcon("rnorvegicus", "hsapiens")
df.human2human <- orthIDsingle("hsapiens")

set.seed(1)
list.allorthologs <- list(
  mmusculus = df.mouse2human[sample(nrow(df.mouse2human), 100)], 
  rnorvegicus = df.rat2human[sample(nrow(df.rat2human), 100)], 
  hsapiens = df.human2human[sample(nrow(df.human2human), 100)])

df.allorthologs <- do.call(rbind, list.allorthologs)

# from those human genes with multiple mappings, 
# pick one at random
df.orth <- df.allorthologs %>%
  select(gene, ortholog) %>%
  group_by(ortholog) %>%
  summarise_all(function(x) head(x, 1))
```

## Human data

```{r}
set.seed(2)
id.signal <- sample(df.orth$ortholog, 150)
id.noise <- setdiff(df.orth$ortholog, id.signal)

df.signal <- filter(df.orth, ortholog %in% id.signal)
df.noise <- filter(df.orth, ortholog %in% id.noise)

n.hsa.pergroup <- 8
k.hsa <- nrow(df.orth)
vals.hsa.y <- c(-2, -1, 0, 1, 2)
means.hsa.x <- c(-2, -1, 0, 1, 2)

y <- rep(vals.hsa.y, times = n.hsa.pergroup)

generate.mat <- function(gene.ids, n, prefix = "sample", seed = 1, ...) {
    set.seed(seed)
    k <- length(gene.ids)
    mat <- matrix(rnorm(n = k*n, ...), ncol = k)
    colnames(mat) <- gene.ids
    rownames(mat) <- paste0(prefix, seq_len(n))
    mat
}
generate.fc <- function(gene.ids, seed = 1, ...) {
  set.seed(seed)
  
  df <- data.frame(
    logFC = rnorm(n = length(gene.ids), ...), 
    adj.P.Val = runif(n = length(gene.ids))
  )
  rownames(df) <- gene.ids
  
  df
}

# generate signal
list.X.signal <- mapply(
    generate.mat, 
    mean = means.hsa.x, 
    seed = seq_along(means.hsa.x), 
    prefix = paste0("x_y_", means.hsa.x, "_"), 
    MoreArgs = list(
      gene.ids = id.signal, 
      n = n.hsa.pergroup, 
      sd = 2
    ), 
    SIMPLIFY = FALSE
)

# generate noise
list.X.noise <- mapply(
    generate.mat, 
    prefix = paste0("x_y_", means.hsa.x, "_"), 
    seed = seq_along(means.hsa.x), 
    MoreArgs = list(
      gene.ids = id.noise, 
      n = n.hsa.pergroup, 
      mean = 0, 
      sd = 2
    ), 
    SIMPLIFY = FALSE
)

X.hsa <- cbind(
  do.call(rbind, list.X.signal), 
  do.call(rbind, list.X.noise)
)


y.hsa <- rep(vals.hsa.y, each = n.hsa.pergroup) %>%
    set_names(rownames(X.hsa)) 


de.genes <- IST::get.de.genes(X.hsa, y.hsa)
```

## Murine FCs

```{r}
set.seed(3)

fc.mmu.rnd1 <- generate.fc(intersect(df.signal$gene, df.mouse2human$gene), seed = 4)
fc.mmu.rnd2 <- generate.fc(intersect(df.noise$gene, df.mouse2human$gene), seed = 5)

fc.rno.true1 <- generate.fc(intersect(df.signal$gene, df.rat2human$gene), seed = 10)
fc.rno.true1$logFC <- 1

fc.rno.true2 <- head(fc.rno.true1, nrow(fc.rno.true1)/2)

fc.hsa.true1 <- generate.fc(df.signal$ortholog, seed = 6, mean = 1, sd = .15)
fc.hsa.opposite1 <- generate.fc(df.orth$ortholog, seed = 7, mean = -3, sd = .5)
```

```{r}
list.sig <- list(
  mmu.rnd.1 = fc.mmu.rnd1, 
  mmu.rnd.2 = fc.mmu.rnd2, 
  rno.true.1 = fc.rno.true1, 
  rno.true.2 = fc.rno.true2, 
  hsa.true.1 = fc.hsa.true1, 
  hsa.opposite.1 = fc.hsa.opposite1
)

# ..and dataframe and metadata
set.seed(1)
df.meta.sig <- data.frame(
  sig.id = names(list.sig), 
  sig.name = c("Mouse random AnMod", "Mouse random Trt", 
               "Rat true AnMod", "Rat true AnMod (half)", 
               "Human true AnMod", "Human overshoot Trt"),
  sig.org = c("mmusculus", "mmusculus", 
              "rnorvegicus", "rnorvegicus", 
              "hsapiens", "hsapiens"), 
  sig.source = c("Study1", "Study1", "Study1", 
                 "Study1", "Study2", "Study3"), 
  sig.type = c("AnimalModel", "Treatment", "AnimalModel", 
               "AnimalModel", "AnimalModel", "Treatment"), 
  sig.db = rep("Made-up", 6), 
  sig.hassignal = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE), 
  stringsAsFactors = FALSE
)
```

```{r}
sapply(list.sig, nrow)
```

## Pathway data

```{r}
generate_pathways <- function(genes, sizes, prefix = "path", seed = 1) {
  set.seed(seed)
  ls.path <- setNames(sizes, paste0(prefix, seq_along(sizes)))
  
  plyr::ldply(
    ls.path, function(x) 
      data.frame(gene.id = sample(genes, size = x)), 
    .id = "path.id"
  ) %>% mutate_all(as.character)
}

path.hsa.signal <- generate_pathways(id.signal, c(75, 25), prefix = "signal", seed = 1)
path.hsa.mixed <- generate_pathways(df.orth$ortholog, c(200, 50), prefix = "mixed", seed = 2) 
path.hsa.noise <- generate_pathways(id.noise, c(75, 25), prefix = "noise", seed = 1)

df.path <- rbind(
  path.hsa.signal, 
  path.hsa.mixed, 
  path.hsa.noise
)

df.meta.path <- data.frame(
  path.id = c("signal1", "signal2", "mixed1", "mixed2", "noise1", "noise2"), 
  path.name = c("Signal large", "Signal small", 
                "Mixed large", "Mixed small", 
                "Noisy large", "Noisy small"),
  path.source = c("[A]", "[A]", "[A]", 
                  "[C]", "[C]", "[C]"), 
  path.type = c("Type1", "Type2", "Type2", 
                "Type3", "Type3", "Type4"), 
  path.db = "Made-up", 
  path.hassignal = c(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE), 
  stringsAsFactors = FALSE
)
```

## Descriptive 

### Reference data

Low dimensional representation

```{r}
pca.hsa <- pca(X.hsa, nPcs = 2, scale = "uv", center = TRUE)

df.pca <- cbind(scores(pca.hsa), data.frame(y = as.factor(y.hsa)))

ggplot(df.pca, aes(x = PC1, y = PC2, colour = y)) +
  geom_point()
```

### Signatures

Distributions of logFCs

```{r}
ldply(list.sig, identity, .id = "name") %>%
  ggplot(aes(x = logFC)) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_histogram() +
  facet_wrap(~name)
```

### Pathways

Size and proportion of truly differential genes

```{r}
group_by(df.path, path.id) %>%
  summarise(n_genes = n(), 
            prop_true = mean(gene.id %in% id.signal))
```


## Save data

```{r}
sample.env.ist <- list(
    X.hsa = X.hsa, 
    y.hsa = y.hsa, 
    de.genes = de.genes, 
    list.sig = list.sig, 
    df.meta.sig = df.meta.sig, 
    list.orth = list.allorthologs, 
    df.path = df.path, 
    df.meta.path = df.meta.path
) %>% list2env

save(sample.env.ist, file = "../../data/sample.env.ist.RData", compress = "xz")
```

## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
