---
title: "Built-in orthology mappings"
author: "Sergio Picart-Armada"
date: "24th March, 2020"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      error = FALSE, warning = FALSE, 
                      fig.height = 4)
```

## Loading packages

```{r}
library(tidyverse)
library(data.table)

if (!dir.exists("03_data")) dir.create("03_data")

devtools::load_all()

set.seed(1)
```

For reproducibility

```{r}
# https://useast.ensembl.org/info/website/archives/index.html
url.current <- "jan2020.archive.ensembl.org"
```


## Orthology mapping

```{r}
# this uses IST's orthIDcon function
v.orgfrom <- c("mmusculus", "rnorvegicus", "hsapiens")
data.list.orth <- lapply(
  setNames(v.orgfrom, v.orgfrom), 
  orthIDcon, 
  org.to = "hsapiens", 
  host = url.current)
comment(data.list.orth) <- paste0(
  "Date: ", date(), 
  " || biomaRt version: ", packageVersion("biomaRt"), 
  " || Ensembl archive: ", url.current)
```

Number of mappings

```{r}
sapply(data.list.orth, nrow)
```

How these look like

```{r}
lapply(data.list.orth, head)
```

Summaries:

```{r}
lapply(data.list.orth, summary)
```

Save internal file

```{r}
save(data.list.orth, file = "../data/data.list.orth.RData", compress = "xz")
```


## Gene labels

Provide by default a data.frame mapping unique ENSEMBL gene identifiers to 
more readable labels.

Original mapping:

```{r}
ensembl.hsa <- biomaRt::useMart(
    host = url.current, 
    biomart = "ensembl", 
    dataset = "hsapiens_gene_ensembl")

map.hsa <- biomaRt::getBM(
    attributes = c("entrezgene_id", "ensembl_gene_id", "external_gene_name"),
    values = TRUE,
    mart = ensembl.hsa) %>%
  mutate(entrezgene_id = as.character(entrezgene_id), 
         external_gene_name = ifelse(external_gene_name == "", NA, external_gene_name))
```

Multiple entrez/symbols are collapsed using `||` as separator.
The ENSEMBL id is not only appended for traceability,
but because there can be several ENSEMBL identifiers mapping to the 
same entrez/symbols, causing collisions in plots.

```{r}
data.dt.genelabels <- data.table(map.hsa)

# use "-" to identify NAs
data.dt.genelabels[is.na(data.dt.genelabels)] <- "-"

data.dt.genelabels <- data.dt.genelabels[
  , by = ensembl_gene_id, 
  lapply(.SD, function(x) paste(unique(x), collapse = "||")), 
  .SDcols = c("entrezgene_id", "external_gene_name")]

# paste ensembl id at the end
data.dt.genelabels[
  , ':='(
    entrezgene_id = paste0(entrezgene_id, " (", ensembl_gene_id, ")"), 
    external_gene_name = paste0(external_gene_name, " (", ensembl_gene_id, ")"), 
    # Update: added a posteriori to have succint labels, only ENSG when necessary
    external_gene_name_only = ifelse(is.na(external_gene_name), paste0("- (", ensembl_gene_id, ")"), external_gene_name)
  )
]

# Update: duplicates appear when using external_gene as a main label
# for those that have a duplicate, group by gene name, and add the ENSG
v.external.gene.duplicated <- subset(data.dt.genelabels, duplicated(external_gene_name_only))$external_gene_name_only %>%
    unique

data.dt.genelabels[
    external_gene_name_only %in% v.external.gene.duplicated, by = external_gene_name_only, 
    external_gene_name_only_ensg := paste0(external_gene_name_only, " (", ensembl_gene_id, ")")
]

# Update: those that are unique, just keep the name. 
# Remove the column with name only (contains duplicates!)
data.dt.genelabels[is.na(external_gene_name_only_ensg), external_gene_name_only_ensg := external_gene_name_only] 
data.dt.genelabels[, external_gene_name_only := NULL]
```

Dimensions of the original mapping and that without duplicate ENSEMBL ids

```{r}
dim(map.hsa)
dim(data.dt.genelabels)
```

Looks of the mapping:

```{r}
head(data.dt.genelabels)
```

Genes with multiple symbols:

```{r}
data.dt.genelabels[grepl("\\|\\|", external_gene_name)]
```

Genes with multiple entrez ids:

```{r}
data.dt.genelabels[grepl("\\|\\|", entrezgene_id)]
```

Make sure there are no missings or duplicates

```{r}
stopifnot(sum(is.na(data.dt.genelabels)) == 0)

v.dup <- colSums(apply(data.dt.genelabels, 2, duplicated))
stopifnot(all(v.dup == 0))
```

Provide as named vectors ready to use

```{r}
vec.ensembl2symbol <- setNames(
  data.dt.genelabels$external_gene_name, data.dt.genelabels$ensembl_gene_id)
head(vec.ensembl2symbol)

vec.ensembl2entrez <- setNames(
  data.dt.genelabels$entrezgene_id, data.dt.genelabels$ensembl_gene_id)
head(vec.ensembl2entrez)

# Update: succint labels
vec.ensembl2symbolonly <- setNames(
  data.dt.genelabels$external_gene_name_only_ensg, data.dt.genelabels$ensembl_gene_id)
head(vec.ensembl2symbolonly)

```

Save internal file

```{r}
save(vec.ensembl2symbol, file = "../data/vec.ensembl2symbol.RData", compress = "xz")
save(vec.ensembl2symbolonly, file = "../data/vec.ensembl2symbolonly.RData", compress = "xz")
save(vec.ensembl2entrez, file = "../data/vec.ensembl2entrez.RData", compress = "xz")
```

## Update: Gene labels with improved readability

Idea: for gene plots, space gets cluttered with ENSG labels, mostly useful for traceability anyway. 
Plots can be generated with and without the verbose labels. 
The only concern is to make sure we have as little mismatchings as possible (ENGS without a symbol)

```{r}
# map.hsa <- biomaRt::getBM(
#     attributes = c("entrezgene_id", "ensembl_gene_id", "external_gene_name", "hgnc_symbol"),
#     values = TRUE,
#     mart = ensembl.hsa) %>%
#   mutate(entrezgene_id = as.character(entrezgene_id), 
#          hgnc_symbol = ifelse(hgnc_symbol == "", NA, hgnc_symbol), 
#          external_gene_name = ifelse(external_gene_name == "", NA, external_gene_name))
```

Tried different attributes, the one that gives the most mappings is `external_gene_name`.
Tested and added to the code above.


## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
